
```{r}
#Main loader block
library(pheatmap)
library(RColorBrewer)
library(dplyr)
library(tidyr)
library(ontologyIndex)
library(tibble)
library(ggplot2)
library(ggtext)
library(ggrepel)
library(readxl)
library(stringr)
library(limma)

source("util_functions.R", local=TRUE)$value

```

```{r}
expression <- read_excel("B.subtilis_expression.xlsx")
expression <- as.data.frame(expression)

```

```{r}
excel_data <- read_excel("2025-05-27_12-52-12_subti_wiki_export_genes.xlsx")
excel_data$id <- as.character(excel_data$id)

# In excel_data: create a normalized key that only removes "_" for BUS*
excel_data <- excel_data %>%
  mutate(
    locus_norm = if_else(
      str_detect(as.character(`genomic_annotation.locus_tag`), "^BSU_[0-9]+$"),
      str_replace_all(as.character(`genomic_annotation.locus_tag`), "_", ""),
      as.character(`genomic_annotation.locus_tag`)
    ))

# Join to get the 'id' from excel_data to expression by matching the keys
expression_tmp <- expression %>%
  left_join(
    excel_data %>% select(locus_norm, id),
    by = c("Locus_tag" = "locus_norm")
  ) %>%
  mutate(id = as.character(id))
# Convert 'id' to row names and remove the column
expression_df <- expression_tmp %>%
  column_to_rownames(var = "id") %>%
  as.data.frame(check.names = FALSE)

expression <- expression_df[ ,-c(1:12)]
# convert values to numbers
expression[] <- lapply(expression, as.numeric)

dim(expression)

```
# filtering out non-coding genes and genes with low expression levels
```{r}
genes_data <- read_excel("2025-05-27_12-52-12_subti_wiki_export_genes.xlsx", sheet = "Export")
coding_genes <- genes_data %>%
  filter(!is.na(genomic_annotation.prot_sequence))

filtered_expression_data <- expression[rownames(expression) %in% coding_genes$id, ]
dim(filtered_expression_data)


# Filter out genes that don't have an FPKM above 10 in at least 10% of the samples
num_samples_above_threshold <- rowSums(filtered_expression_data > 10)
min_samples <- 0.1 * ncol(filtered_expression_data)
expression_data_clean <- filtered_expression_data[num_samples_above_threshold >= min_samples, ]
dim(expression_data_clean)

```

#Assessment of Sample Variation Prior to Clustering
```{r}
expression_data <- expression_data_clean[,c(1:150)]

# Compute correlation between samples
sample_cor_filtered <- cor(expression_data, method = 'spearman')

# Define a fixed color scale
breaks <- seq(-1, 1, length.out = 200)
colors <- colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(length(breaks) - 1)
# Heatmap
png("heatmap.png", width = 3000, height = 4000, res = 400)
pheatmap::pheatmap(as.matrix(sample_cor_filtered),
                   clustering_rows = FALSE,
                   clustering_cols = FALSE,
                   treeheight_row = 0,
                   treeheight_col = 0,
                   main = "Heatmap of the sample-to-sample correlations: B. subtilis",
                   show_colnames = FALSE,
                   show_rownames = FALSE,
                   color = colors,
                   breaks = breaks,
                   border_color = rgb(0, 0, 0, alpha = 0.1)
                   )
dev.off()
```

Reading the GO annotation file from EggNOG and saving a cleaned version
```{r}
# Read the Excel file, skipping metadata rows
df <- read_excel("out.emapper.annotations_B.subtilis.xlsx", skip = 2)

# Extract gene IDs and GO annotations (1st and 10th columns)
gene_go_df <- df[, c(1, 10)]
colnames(gene_go_df) <- c("Gene", "annotation")

gene_go_df <- gene_go_df[!is.na(gene_go_df$annotation) & gene_go_df$annotation != "-", ]
gene_go_df$annotation <- gsub(",", " ", gene_go_df$annotation)
write.table(gene_go_df, "GO_B.subtilis.tab", sep = "\t", row.names = FALSE, quote = FALSE)
```


# Clustering Pipeline (Hierarchical)
```{r}
# === Define input data ===
# Load expression data
expression_data <- expression_data_clean
# GO annotation data
go_annotation <- "GO_B.subtilis.tab"

# === Define cutoff points, transformations, and correlation methods ===
cutoff_points <- seq(0.1, 1, by = 0.05)
user_transformations <- c("none", "sqrt")  # or c("none", "sqrt", "log")
correlation_methods <- c("pearson", "spearman")
# ======

# Load GO data and parse the GO terms
go_data <- read.table(go_annotation, header = TRUE, sep = "\t", stringsAsFactors = FALSE)
go_data_long <- go_data %>%
  dplyr::select(Gene = Gene, GO_Terms = annotation) %>%
  separate_rows(GO_Terms, sep = " ")
go_data_long$Gene <- as.character(go_data_long$Gene)

all_genes <- row.names(expression_data) #including all genes for enrichment analysis, regardless of having GO annotation or not

# Initialize result containers
all_enrichment_tables <- list()
results <- data.frame()
cluster_assignments <- list()

# Main loop over configurations
for (correlation_method in correlation_methods) {
  if (correlation_method == "spearman") {
    transformation_names <- "none"
  } else if (correlation_method == "pearson") {
    transformation_names <- user_transformations
  }
  for (transformation_name in transformation_names) {
    for (cutoff_point in cutoff_points) {
      configuration <- paste("hierarchical", transformation_name, correlation_method, cutoff_point, sep = "_")
      cat(sprintf("Processing: %s\n", configuration))
      
      # Applying transformation, and transposing the data for computing gene‑gene correlations
      expression_data_t <- transformations[[transformation_name]](t(expression_data))
      # Correlation and clustering
      gene_cor <- cor(expression_data_t, method = correlation_method)
      gene_dist <- as.dist(1 - gene_cor)
      gene_hc <- hclust(gene_dist)
      clusters <- cutree(gene_hc, h = cutoff_point)
      
      cluster_table <- data.frame(Gene = names(clusters), Cluster = as.vector(clusters))
      #save cluster assignment for this configuration
      cluster_assignments[[configuration]] <- cluster_table

      # ✅ Call the Function to test the enrichment of a given clustering
      out <- test_cluster_enrichment(cluster_table, go_data_long, configuration)
      
      # Append summary and full enrichment
      results <- rbind(results, out$summary)
      all_enrichment_tables[[length(all_enrichment_tables) + 1]] <- out$enrichment
      
    }
  }
}

# === Combine All Enrichment Tables ===
results_df <- as.data.frame(results)

enrichment_tables <- dplyr::bind_rows(all_enrichment_tables)
row.names(enrichment_tables) <- NULL
# ===

# Compute product (objective score)
results_df$product <- ifelse(is.na(results_df$Mean_odds), 0,
                             results_df$Mean_odds * results_df$Total_Clusters_With_More_Than_5_Genes)

# find the max product and return its clustering
best_idx     <- which.max(results_df$product)
best_config  <- results_df$Configuration[best_idx]
best_product <- results_df$product[best_idx]
cat(sprintf("Best configuration: %s (product = %.4f)\n", best_config, best_product))

# Retrieve the stored gene→cluster assignment for the best configuration
best_cluster_assignment <- cluster_assignments[[best_config]]
```

hierarchical visualization
```{r}
# product (Clustering Quality)
results_parsed <- results_df %>%
  separate(Configuration, into = c("Method", "Transformation", "Correlation", "Cutoff_Point"), sep = "_", fill = "right")

max_row <- results_parsed[which.max(results_parsed$product), ]
max_row$label <- paste0("Max: ", round(max_row$product, 2),
                        " / Tf: ", max_row$Transformation,
                        " / COR: ", max_row$Correlation,
                        " / Cut: ", max_row$Cutoff_Point)

results_parsed$Cutoff_Point <- as.numeric(results_parsed$Cutoff_Point)
max_row$Cutoff_Point <- as.numeric(max_row$Cutoff_Point)

p <- ggplot(results_parsed, aes(x = Cutoff_Point, y = product,
                                color = Correlation, shape = Transformation)) +
  geom_point(data = max_row, aes(x = Cutoff_Point, y = product),
             inherit.aes = FALSE, color = "black", size = 3, shape = 8) +
  geom_text_repel(data = max_row, aes(x = Cutoff_Point, y = product, label = label), inherit.aes = FALSE, color = "black", size = 3, fontface = "bold", nudge_y = 1000) +
  geom_line(aes(group = interaction(Correlation, Transformation))) +
  geom_point(size = 3) +
  scale_shape_manual(values = c(0, 1, 2))  +
  labs(
    x = "**Cutoff Point**",
    y = "**Clustering Quality**",
    color = "Correlation",
    shape = "Transformation"
  ) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal() +
  scale_x_continuous(breaks = seq(0.1, 1, by = 0.1)) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.text  = element_text(face = "bold"),
    axis.title.x = element_markdown(),
    axis.title.y = element_markdown()
  )
ggsave("Cutoff_vs_Product.png", plot = p, width = 6, height = 4, dpi = 300)


# mean odds ratio (Strength of Enrichment)
max_or <- results_parsed[which.max(results_parsed$Mean_odds), ]
max_or$label <- paste0("Max: ", round(max_or$Mean_odds, 2),
                        " / Tf: ", max_or$Transformation,
                        " / COR: ", max_or$Correlation,
                        " / Cut: ", max_or$Cutoff_Point)

OR <- ggplot(results_parsed, aes(x = Cutoff_Point, y = Mean_odds,
                                color = Correlation, shape = Transformation)) +
  geom_point(data = max_or, aes(x = Cutoff_Point, y = Mean_odds),
             inherit.aes = FALSE, color = "black", size = 3, shape = 8) +
  geom_text_repel(data = max_or, aes(x = Cutoff_Point, y = Mean_odds, label = label), inherit.aes = FALSE, color = "black", size = 3, fontface = "bold", nudge_y = 10) +
  geom_line(aes(group = interaction(Correlation, Transformation))) +
  geom_point(size = 3) +
  scale_shape_manual(values = c(0, 1, 2))  +
  labs(
    x = "**Cutoff Point**",
    y = "**Strength of Enrichment**",
    color = "Correlation",
    shape = "Transformation"
  ) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal() +
  scale_x_continuous(breaks = seq(0.1, 1, by = 0.1)) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.text  = element_text(face = "bold"),
    axis.title.x = element_markdown(),
    axis.title.y = element_markdown()
  )
ggsave("Cutoff_vs_OR.png", plot = OR, width = 5.7, height = 3.7, dpi = 300)


# clusters with ≥5 genes
max_clusters <- results_parsed[which.max(results_parsed$Total_Clusters_With_More_Than_5_Genes), ]
max_clusters$label <- paste0("Max: ", max_clusters$Total_Clusters_With_More_Than_5_Genes,
                        " / Tf: ", max_clusters$Transformation,
                        " / COR: ", max_clusters$Correlation,
                        " / Cut: ", max_clusters$Cutoff_Point)

C <- ggplot(results_parsed, aes(x = Cutoff_Point, y = Total_Clusters_With_More_Than_5_Genes,
                                color = Correlation, shape = Transformation)) +
  geom_point(data = max_clusters, aes(x = Cutoff_Point, y = Total_Clusters_With_More_Than_5_Genes), inherit.aes = FALSE, color = "black", size = 3, shape = 8) +
  geom_text_repel(data = max_clusters, aes(x = Cutoff_Point, y = Total_Clusters_With_More_Than_5_Genes, label = label), inherit.aes = FALSE, color = "black", size = 3, fontface = "bold", nudge_y = 25) +
  geom_line(aes(group = interaction(Correlation, Transformation))) +
  geom_point(size = 3) +
  scale_shape_manual(values = c(0, 1, 2))  +
  labs(
    x = "**Cutoff Point**",
    y = "**Clusters ≥ 5**",
    color = "Correlation",
    shape = "Transformation"
  ) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal() +
  scale_x_continuous(breaks = seq(0.1, 1, by = 0.1)) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.text  = element_text(face = "bold"),
    axis.title.x = element_markdown(),
    axis.title.y = element_markdown()
  )
ggsave("Cutoff_vs_Cluster.png", plot = C, width = 5.7, height = 3.7, dpi = 300)

```


# Differential Expression
clustering with optimal condition
```{r}
expression_data_clean_t <- t(expression_data_clean)
#expression_data_clean_t <- log(t(expression_data_clean) + 1)
gene_cor <- cor(expression_data_clean_t, method = "pearson")
gene_dist <- as.dist(1 - gene_cor)
gene_hc <- hclust(gene_dist)
clusters <- cutree(gene_hc, h = 0.25)
cluster_table <- data.frame(Gene = names(clusters), Cluster = clusters)

cluster_sizes <- table(cluster_table$Cluster)
valid_clusters <- names(cluster_sizes[cluster_sizes >= 5])
cluster_table <- subset(cluster_table, Cluster %in% valid_clusters)

```

Differential Expression for Microarray: Glu vs Fru (limma)
```{r}
# Sample columns of interest:
glu_samples <- c("Glu_1_hyb15403702","Glu_2_hyb15349202","Glu_3_hyb15349002")
fru_samples <- c("Fru_1_hyb15354002","Fru_2_hyb15345702","Fru_3_hyb15352002")

expression_filtered <- expression[, c(glu_samples, fru_samples)]

stopifnot(all(c(glu_samples, fru_samples) %in% colnames(expression_filtered)))
expr_mat <- as.matrix(expression_filtered)
mode(expr_mat) <- "numeric"

# Optional: between-array normalization (quantile). # Since data are already log2
expr_norm <- normalizeBetweenArrays(expr_mat, method = "quantile")

# 2) Design and contrasts
group <- factor(c(rep("Glu", length(glu_samples)),
                  rep("Fru", length(fru_samples))),
                levels = c("Glu","Fru"))
design <- model.matrix(~ 0 + group)
colnames(design) <- c("Glu","Fru")
contrast.matrix <- makeContrasts(Fru_vs_Glu = Fru - Glu, levels = design)

# 3) Fit limma model
fit  <- lmFit(expr_norm, design)
fit2 <- contrasts.fit(fit, contrast.matrix)
fit2 <- eBayes(fit2)

# 4) Results table
res <- topTable(fit2, coef = "Fru_vs_Glu", number = Inf, adjust.method = "BH", sort.by = "P")
res$Gene <- rownames(res)
res <- res[, c("Gene","logFC","AveExpr","t","P.Value","adj.P.Val","B")]
```

differential expression plot
```{r}
# 1) Clean and classify results
results <- res %>%
  filter(!is.na(adj.P.Val), !is.na(logFC)) %>%
  mutate(
    regulation_class = case_when(
      logFC >  1 ~ "Upregulated (log2FC > 1)",
      logFC < -1 ~ "Downregulated (log2FC < -1)",
      TRUE                ~ "|log2FC| ≤ 1"
      ),
    alpha_class = if_else(adj.P.Val < 0.05, "padj < 0.05 (solid)", "padj ≥ 0.05 (transparent)")
    )

# 2) Merge with clusters
results_cluster <- cluster_table %>%
  inner_join(results, by = "Gene") %>%
  arrange(Cluster)

results_cluster$Cluster <- factor(
  results_cluster$Cluster,
  levels = unique(results_cluster$Cluster)
  )

results_cluster$ClusterGroup <- cut(
  as.numeric(factor(results_cluster$Cluster)),
  breaks = 2
)

# 3)plot
p <- ggplot(results_cluster, aes(x = Cluster, y = logFC)) +
  geom_point(
  aes(color = regulation_class, alpha = alpha_class),
  size = 2,
  stroke = 0
  ) +
  scale_color_manual(
    values = c(
      "Upregulated (log2FC > 1)"    = "red",
      "Downregulated (log2FC < -1)" = "blue",
      "|log2FC| ≤ 1"                = "grey60"
      ),
    guide = guide_legend(nrow = 1)
    ) +
  scale_alpha_manual(
    values = c("padj < 0.05 (solid)" = 1, "padj ≥ 0.05 (transparent)" = 0.3),
    guide = guide_legend(nrow = 1)
    ) +
  labs(
    title = "Differential Expression Analysis in B. subtilis: Fructose vs Glucose growth",
    color = "Regulation",
    alpha  = "padj",
    y = "log2FoldChange"
    ) +
  facet_wrap(~ ClusterGroup, scales = "free_x", ncol = 1) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text = element_blank(),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
    legend.position = "bottom",
    legend.title = element_text(size = 10),
    legend.margin = margin(t = 6)
    ) +
  scale_y_continuous(limits = c(-3, 5), breaks = seq(-20, 20, by = 2))+
  theme(
  #axis.text.x = element_blank()
  )+
  geom_hline(yintercept = 0, linewidth = 0.1, color = "grey30")+
  theme(
  panel.spacing = unit(1.2, "lines")
  )

ggsave("my_facet_plot.png", plot = p, width = 9.5, height = 6, dpi = 500)
```

