

```{r}
#Main loader block
library(DESeq2)
library(pheatmap)
library(RColorBrewer)
library(dplyr)
library(tidyr)
library(ontologyIndex)
library(tibble)
library(ggplot2)
library(ggtext)
library(ggrepel)
library(tibble)
library(readxl)

source("util_functions.R", local=TRUE)$value

```


Removing samples with SRRxxx identifiers from the counts matrix. These samples are excluded from the clustering analysis but retained for downstream differential expression analysis once cluster assignments have been determined.
```{r}
data <- read.table('C1_ATCC_counts.tab',header=T,row.names=1,sep='\t',check.names=F,quote='')
data <- data[, 1:89]

```

Computing normalized counts using DESeq and converting them to FPKM values. Clustering is performed on the FPKM matrix
```{r}
mm <- data.frame(row.names=colnames(data),condition=colnames(data),type="paired-end")
dds <- DESeqDataSetFromMatrix(countData=data,
                                  colData=mm,
                                  design=~condition)
dds <- estimateSizeFactors(dds)

genes <- read.table('C1_ATCC_genes.tab',header=T,row.names=1,sep='\t',check.names=F,quote='')
genes = genes[rownames(data),,F]
if ('Length' %in% colnames(genes)) {
genes$Start = 1
genes$End = genes$Start + genes$Length
genes$Strand = '*'
#genes$chr = 'chrom'
} 
genesGR <- GRanges(genes)
rowRanges(dds) <- genesGR
res_fpkm <- fpkm(dds)
write.table(res_fpkm,'fpkm_counts.tab',quote=F,sep='\t')
```

#Loading the FPKM expression matrix and filtering out genes with low expression levels
```{r}
fpkm_counts <- read.table("fpkm_counts.tab", header = TRUE, sep = "\t", row.names = 1, check.names=F)
dim(fpkm_counts)

# Calculate the number of samples where the FPKM value is above the threshold for each gene
num_samples_above_threshold <- rowSums(fpkm_counts > 10)
# Filter out genes that don't have an FPKM above threshold in at least 20% of the samples
min_samples <- 0.2 * ncol(fpkm_counts)
fpkm_counts_filtered <- fpkm_counts[num_samples_above_threshold >= min_samples, ]

dim(fpkm_counts_filtered)
```

#Assessment of Sample Variation Prior to Clustering
```{r}
# Compute correlation between samples
sample_cor_filtered <- cor(fpkm_counts_filtered, method = 'spearman')

# Define a fixed color scale
breaks <- seq(-1, 1, length.out = 300)
colors <- colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(length(breaks) - 1)
# Heatmap
png("heatmap.png", width = 3000, height = 4000, res = 400)
pheatmap::pheatmap(as.matrix(sample_cor_filtered),
         clustering_rows = FALSE,
         clustering_cols = FALSE,
         treeheight_row = 0,
         treeheight_col = 0,
         main = "Heatmap of the sample-to-sample correlations: C1",
         show_colnames = FALSE,
         show_rownames = FALSE,
         color = colors,
         breaks = breaks,
         border_color = rgb(0, 0, 0, alpha = 0.1)
         )
dev.off()
```


Reading the GO annotation file from EggNOG and saving a cleaned version
```{r}
# Read the Excel file, skipping metadata rows
df <- read_excel("out.emapper.annotations_C1.xlsx", skip = 2)

# Extract gene IDs and GO annotations (1st and 10th columns)
gene_go_df <- df[, c(1, 10)]
colnames(gene_go_df) <- c("Gene", "annotation")

gene_go_df <- gene_go_df[!is.na(gene_go_df$annotation) & gene_go_df$annotation != "-", ]
gene_go_df$annotation <- gsub(",", " ", gene_go_df$annotation)
write.table(gene_go_df, "GO_C1.tab", sep = "\t", row.names = FALSE, quote = FALSE)
```


# Clustering Pipeline (Hierarchical)
```{r}
# === Define input data ===
# Load expression data
expression_data <- fpkm_counts_filtered
# GO annotation data
go_annotation <- "GO_C1.tab"

# === Define cutoff points, transformations, and correlation methods ===
cutoff_points <- seq(0.1, 1, by = 0.05)
user_transformations <- c("sqrt")  # or c("none", "sqrt", "log")
correlation_methods <- c("pearson", "spearman")
# ======

# Load GO data and parse the GO terms
go_data <- read.table(go_annotation, header = TRUE, sep = "\t", stringsAsFactors = FALSE)
go_data_long <- go_data %>%
  dplyr::select(Gene = Gene, GO_Terms = annotation) %>%
  separate_rows(GO_Terms, sep = " ")
go_data_long$Gene <- as.character(go_data_long$Gene)

all_genes <- row.names(expression_data) #including all genes for enrichment analysis, regardless of having GO annotation or not

# Initialize result containers
all_enrichment_tables <- list()
results <- data.frame()
cluster_assignments <- list()

# Main loop over configurations
for (correlation_method in correlation_methods) {
  if (correlation_method == "spearman") {
    transformation_names <- "none"
  } else if (correlation_method == "pearson") {
    transformation_names <- user_transformations
  }
  for (transformation_name in transformation_names) {
    for (cutoff_point in cutoff_points) {
      configuration <- paste("hierarchical", transformation_name, correlation_method, cutoff_point, sep = "_")
      cat(sprintf("Processing: %s\n", configuration))
      
      # Applying transformation, and transposing the data for computing gene‑gene correlations
      expression_data_t <- transformations[[transformation_name]](t(expression_data))
      # Correlation and clustering
      gene_cor <- cor(expression_data_t, method = correlation_method)
      gene_dist <- as.dist(1 - gene_cor)
      gene_hc <- hclust(gene_dist)
      clusters <- cutree(gene_hc, h = cutoff_point)
      
      cluster_table <- data.frame(Gene = names(clusters), Cluster = as.vector(clusters))
      #save cluster assignment for this configuration
      cluster_assignments[[configuration]] <- cluster_table

      # ✅ Call the Function to test the enrichment of a given clustering
      out <- test_cluster_enrichment(cluster_table, go_data_long, configuration)
      
      # Append summary and full enrichment
      results <- rbind(results, out$summary)
      all_enrichment_tables[[length(all_enrichment_tables) + 1]] <- out$enrichment
      
    }
  }
}

# === Combine All Enrichment Tables ===
results_df <- as.data.frame(results)

enrichment_tables <- dplyr::bind_rows(all_enrichment_tables)
row.names(enrichment_tables) <- NULL
# ===

# Compute product (objective score)
results_df$product <- ifelse(is.na(results_df$Mean_odds), 0,
                             results_df$Mean_odds * results_df$Total_Clusters_With_More_Than_5_Genes)

# find the max product and return its clustering
best_idx     <- which.max(results_df$product)
best_config  <- results_df$Configuration[best_idx]
best_product <- results_df$product[best_idx]
cat(sprintf("Best configuration: %s (product = %.4f)\n", best_config, best_product))

# Retrieve the stored gene→cluster assignment for the best configuration
best_cluster_assignment <- cluster_assignments[[best_config]]
```

hierarchical visualization
```{r}
# product (Clustering Quality)
results_parsed <- results_df %>%
  separate(Configuration, into = c("Method", "Transformation", "Correlation", "Cutoff_Point"), sep = "_", fill = "right")

max_row <- results_parsed[which.max(results_parsed$product), ]
max_row$label <- paste0("Max: ", round(max_row$product, 2),
                        " / Tf: ", max_row$Transformation,
                        " / COR: ", max_row$Correlation,
                        " / Cut: ", max_row$Cutoff_Point)

results_parsed$Cutoff_Point <- as.numeric(results_parsed$Cutoff_Point)
max_row$Cutoff_Point <- as.numeric(max_row$Cutoff_Point)

p <- ggplot(results_parsed, aes(x = Cutoff_Point, y = product,
                                color = Correlation, shape = Transformation)) +
  geom_point(data = max_row, aes(x = Cutoff_Point, y = product),
             inherit.aes = FALSE, color = "black", size = 3, shape = 8) +
  geom_text_repel(data = max_row, aes(x = Cutoff_Point, y = product, label = label), inherit.aes = FALSE, color = "black", size = 3, fontface = "bold", nudge_y = 1000) +
  geom_line(aes(group = interaction(Correlation, Transformation))) +
  geom_point(size = 3) +
  scale_shape_manual(values = c(0, 1, 2))  +
  labs(
    x = "**Cutoff Point**",
    y = "**Clustering Quality**",
    color = "Correlation",
    shape = "Transformation"
  ) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal() +
  scale_x_continuous(breaks = seq(0.1, 1, by = 0.1)) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.text  = element_text(face = "bold"),
    axis.title.x = element_markdown(),
    axis.title.y = element_markdown()
  )
ggsave("Cutoff_vs_Product.png", plot = p, width = 6, height = 4, dpi = 300)


# mean odds ratio (Strength of Enrichment)
max_or <- results_parsed[which.max(results_parsed$Mean_odds), ]
max_or$label <- paste0("Max: ", round(max_or$Mean_odds, 2),
                        " / Tf: ", max_or$Transformation,
                        " / COR: ", max_or$Correlation,
                        " / Cut: ", max_or$Cutoff_Point)

OR <- ggplot(results_parsed, aes(x = Cutoff_Point, y = Mean_odds,
                                color = Correlation, shape = Transformation)) +
  geom_point(data = max_or, aes(x = Cutoff_Point, y = Mean_odds),
             inherit.aes = FALSE, color = "black", size = 3, shape = 8) +
  geom_text_repel(data = max_or, aes(x = Cutoff_Point, y = Mean_odds, label = label), inherit.aes = FALSE, color = "black", size = 3, fontface = "bold", nudge_y = 10) +
  geom_line(aes(group = interaction(Correlation, Transformation))) +
  geom_point(size = 3) +
  scale_shape_manual(values = c(0, 1, 2))  +
  labs(
    x = "**Cutoff Point**",
    y = "**Strength of Enrichment**",
    color = "Correlation",
    shape = "Transformation"
  ) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal() +
  scale_x_continuous(breaks = seq(0.1, 1, by = 0.1)) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.text  = element_text(face = "bold"),
    axis.title.x = element_markdown(),
    axis.title.y = element_markdown()
  )
ggsave("Cutoff_vs_OR.png", plot = OR, width = 5.7, height = 3.7, dpi = 300)


# clusters with ≥5 genes
max_clusters <- results_parsed[which.max(results_parsed$Total_Clusters_With_More_Than_5_Genes), ]
max_clusters$label <- paste0("Max: ", max_clusters$Total_Clusters_With_More_Than_5_Genes,
                        " / Tf: ", max_clusters$Transformation,
                        " / COR: ", max_clusters$Correlation,
                        " / Cut: ", max_clusters$Cutoff_Point)

C <- ggplot(results_parsed, aes(x = Cutoff_Point, y = Total_Clusters_With_More_Than_5_Genes,
                                color = Correlation, shape = Transformation)) +
  geom_point(data = max_clusters, aes(x = Cutoff_Point, y = Total_Clusters_With_More_Than_5_Genes), inherit.aes = FALSE, color = "black", size = 3, shape = 8) +
  geom_text_repel(data = max_clusters, aes(x = Cutoff_Point, y = Total_Clusters_With_More_Than_5_Genes, label = label), inherit.aes = FALSE, color = "black", size = 3, fontface = "bold", nudge_y = 25) +
  geom_line(aes(group = interaction(Correlation, Transformation))) +
  geom_point(size = 3) +
  scale_shape_manual(values = c(0, 1, 2))  +
  labs(
    x = "**Cutoff Point**",
    y = "**Clusters ≥ 5**",
    color = "Correlation",
    shape = "Transformation"
  ) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal() +
  scale_x_continuous(breaks = seq(0.1, 1, by = 0.1)) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.text  = element_text(face = "bold"),
    axis.title.x = element_markdown(),
    axis.title.y = element_markdown()
  )
ggsave("Cutoff_vs_Cluster.png", plot = C, width = 5.7, height = 3.7, dpi = 300)

```


# GO enrichment results for optimal clustering
```{r}
# Keep only significant rows
enrichment_tables <- enrichment_tables %>% filter(adj_pvalue < 0.05)
# Filter enrichment for the chosen configuration
best_enrichment <- enrichment_tables %>% filter(Configuration == best_config)

# adding GO Description, Ontology, and Definition
go_obo <- get_ontology("go.obo", extract_tags = "everything")
best_enrichment$GO_Term <- trimws(best_enrichment$GO_Term)
# Build reverse lookup for alt_ids -> primary IDs
alt_id_map <- unlist(lapply(names(go_obo$alt_id), function(primary_id) {
  setNames(rep(primary_id, length(go_obo$alt_id[[primary_id]])), go_obo$alt_id[[primary_id]])
}))
# Map GO terms to primary IDs when needed
best_enrichment <- best_enrichment %>%
  mutate(
    GO_Term_Corrected = case_when(
      GO_Term %in% names(go_obo$name) ~ GO_Term,
      GO_Term %in% names(alt_id_map) ~ alt_id_map[GO_Term],
      TRUE ~ NA_character_
    )
  )

available_ids <- unique(na.omit(best_enrichment$GO_Term_Corrected))
go_info <- tibble(
  GO_Term_Corrected = available_ids
) %>%
  distinct(GO_Term_Corrected) %>%
  mutate(
    GO_Description = go_obo$name[GO_Term_Corrected],
    GO_Ontology = go_obo$namespace[GO_Term_Corrected],
    GO_Definition = go_obo$def[GO_Term_Corrected]
  )
# Merge metadata into original table
final_enrichment <- best_enrichment %>%
  left_join(go_info, by = "GO_Term_Corrected")

print(final_enrichment)
```


#differential expression
```{r}
# Load raw counts data
counts <- read.table('C1_ATCC_counts.tab',header=T,row.names=1,sep='\t',check.names=F,quote='')
```

Avicel vs Glucose
```{r}
# Create metadata
metadata <- data.frame(
  row.names = c("SRR30193268", "SRR30193269", "SRR30193270", 
             "SRR30193271", "SRR30193272", "SRR30193273"),
  sugar_source = c("Glucose", "Glucose", "Glucose", 
                   "Avicel", "Avicel", "Avicel"),
  replicate = c(3, 2, 1, 3, 2, 1)
)

filtered_counts <- counts[, colnames(counts) %in% row.names(metadata)]
metadata$sugar_source <- factor(metadata$sugar_source)

# --- DESeq setup ---
# this should be true, rownames of coldata and column names of counts table should match
all(rownames(metadata) %in% colnames(filtered_counts))

# columns of the count matrix must match the rows of the metadata in the same order
all(rownames(metadata) == colnames(filtered_counts))
# If the line above is not TRUE, run this:
# filtered_counts <- filtered_counts[, rownames(metadata)]

dds <- DESeqDataSetFromMatrix(countData = filtered_counts,
                              colData = metadata,
                              design = ~ sugar_source)
dds <- DESeq(dds)

# contrast = c("factorName", "numeratorLevel", "denominatorLevel")
# Get results: Group1 vs Group2 #log2(Group1/Group2)
res <- results(dds, contrast = c("sugar_source", "Avicel", "Glucose"))

DESeq2_results <- as.data.frame(res)

```

day 4 vs day 1
```{r}
metadata <- data.frame(
  row.names = c("H27NLDSX3_104643-001-003_TGAGCTAG-GAACGGTT_L003", "H27WTDSX3_104643-001-003_TGAGCTAG-GAACGGTT_L002",
                "H27NLDSX3_104643-001-006_TTCCAAGG-CTACAAGG_L003", 
             "H27WTDSX3_104643-001-006_TTCCAAGG-CTACAAGG_L002"),
  day = c("D1", "D1",
               "D4", "D4"),
  replicate = c(0, 1, 0, 1)
)

filtered_counts <- counts[, colnames(counts) %in% row.names(metadata)]
metadata$day <- factor(metadata$day)

all(rownames(metadata) %in% colnames(filtered_counts))
all(rownames(metadata) == colnames(filtered_counts))
filtered_counts <- filtered_counts[, rownames(metadata)]


dds <- DESeqDataSetFromMatrix(countData = filtered_counts,
                              colData = metadata,
                              design = ~ day)
dds <- DESeq(dds)

res <- results(dds, contrast = c("day", "D4", "D1"))
DESeq2_results <- as.data.frame(res)
```


Plot for differential expression and cluster visualization
```{r}
# 1) Clean and classify DESeq2_results
DESeq2_results <- DESeq2_results %>%
  filter(!is.na(padj), !is.na(log2FoldChange)) %>%
  mutate(
    regulation_class = case_when(
      log2FoldChange >  1 ~ "Upregulated (log2FC > 1)",
      log2FoldChange < -1 ~ "Downregulated (log2FC < -1)",
      TRUE                ~ "|log2FC| ≤ 1"
    ),
    alpha_class = if_else(padj < 0.05, "padj < 0.05 (solid)", "padj ≥ 0.05 (transparent)")
  ) %>%
  tibble::rownames_to_column("Gene")

# Keep only clusters with more than 4 genes
cluster_sizes <- table(best_cluster_assignment$Cluster)
valid_clusters <- names(cluster_sizes[cluster_sizes >= 5])
cluster_table <- subset(best_cluster_assignment, Cluster %in% valid_clusters)

# 2) Merge with clusters
DESeq2_results_cluster <- cluster_table %>%
  inner_join(DESeq2_results, by = "Gene") %>%
  arrange(Cluster)

DESeq2_results_cluster$Cluster <- factor(
  DESeq2_results_cluster$Cluster,
  levels = unique(DESeq2_results_cluster$Cluster)
)

DESeq2_results_cluster$ClusterGroup <- cut(
  as.numeric(factor(DESeq2_results_cluster$Cluster)),
  breaks = 3
)

# 3)plot
p <- ggplot(subset(DESeq2_results_cluster, baseMean > 50), aes(x = Cluster, y = log2FoldChange)) +
  geom_point(
  aes(color = regulation_class, alpha = alpha_class),
  size = 2,
  stroke = 0
  ) +
  scale_color_manual(
    values = c(
      "Upregulated (log2FC > 1)"    = "red",
      "Downregulated (log2FC < -1)" = "blue",
      "|log2FC| ≤ 1"                = "grey60"
    ),
    guide = guide_legend(nrow = 1)
  ) +
  scale_alpha_manual(
    values = c("padj < 0.05 (solid)" = 1, "padj ≥ 0.05 (transparent)" = 0.3),
    guide = guide_legend(nrow = 1)
  ) +
  labs(
    title = "Differential Expression Analysis in C1: Avicel vs Glucose",
    color = "Regulation",
    alpha  = "padj"
  ) +
  facet_wrap(~ ClusterGroup, scales = "free_x", ncol = 1) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text = element_blank(),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
    legend.position = "bottom",
    legend.title = element_text(size = 10),
    legend.margin = margin(t = 6)
  ) +
  scale_y_continuous(breaks = seq(-20, 20, by = 2))+
  geom_hline(yintercept = 0, linewidth = 0.1, color = "grey30")

ggsave("my_facet_plot.png", plot = p, width = 11, height = 9.5, dpi = 500)
```


#k-means (optional)
```{r eval=FALSE, include=FALSE}
# === Define input data ===
# Load expression data
expression_data <-  fpkm_counts_filtered
# GO annotation data
go_annotation <-   "GO_C1.tab"

# === Define parameters ===
k_values <- seq(100, 4000, by = 200)
# Log-transform the FPKM matrix
log_expression <- log(expression_data + 1)
# ======

# Load GO data and parse the GO terms
go_data <- read.table(go_annotation, header = TRUE, sep = "\t", stringsAsFactors = FALSE)
go_data_long <- go_data %>%
  dplyr::select(Gene = Gene, GO_Terms = annotation) %>%
  separate_rows(GO_Terms, sep = " ")
go_data_long$Gene <- as.character(go_data_long$Gene)

all_genes <- row.names(expression_data) # including all genes for enrichment analysis, regardless of having GO annotation or not

# Initialize result containers
all_enrichment_tables <- list()
results <- data.frame()

# Main loop
for (k in k_values) {
  configuration <- paste("kmeans", k, sep = "_")
  cat(sprintf("Processing: %s\n", configuration))
  
  # Apply k-means clustering
  set.seed(777)  # For reproducibility
  kmeans_result <- kmeans(log_expression, centers = k , nstart = 50, iter.max = 100)
  # Cluster assignments
  clusters_kmeans <- kmeans_result$cluster
  cluster_table <- data.frame(Gene = rownames(log_expression), Cluster = as.vector(clusters_kmeans))
  
  # ✅ Call the Function to test the enrichment of a given clustering
  out <- test_cluster_enrichment(cluster_table, go_data_long, configuration)
      
  # Append summary and full enrichment
  results <- rbind(results, out$summary)
  all_enrichment_tables[[length(all_enrichment_tables) + 1]] <- out$enrichment
}

# === Combine All Enrichment Tables ===
results_kmeans <- as.data.frame(results)
enrichment_tables_k <- dplyr::bind_rows(all_enrichment_tables)
row.names(enrichment_tables_k) <- NULL
```
